---
title: "SAA Analysis LB data"
author: "Gregg Griffenhagen"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE)
library(lme4)
library(lmerTest)
library(tidyverse)
library(magrittr)
library(emmeans)
library(kableExtra)
library(ggrepel)
library(outliers)
library(dplyr)

summarySE <- function(data=NULL, measurevar, groupvars=NULL, na.rm=FALSE,
                      conf.interval=.95, .drop=TRUE) {
  library(plyr)
  library(dplyr)
  
  # New version of length which can handle NA's: if na.rm==T, don't count them
  length2 <- function (x, na.rm=FALSE) {
    if (na.rm) sum(!is.na(x))
    else       length(x)
  }
  
  # This does the summary. For each group's data frame, return a vector with
  # N, mean, and sd
  datac <- ddply(data, groupvars, .drop=.drop,
                 .fun = function(xx, col) {
                   c(N    = length2(xx[[col]], na.rm=na.rm),
                     mean = mean   (xx[[col]], na.rm=na.rm),
                     sd   = sd     (xx[[col]], na.rm=na.rm)
                   )
                 },
                 measurevar
  )
  
  # Rename the "mean" column    
  datac <- rename(datac, c("mean" = measurevar))
  
  datac$se <- datac$sd / sqrt(datac$N)  # Calculate standard error of the mean
  
  # Confidence interval multiplier for standard error
  # Calculate t-statistic for confidence interval: 
  # e.g., if conf.interval is .95, use .975 (above/below), and use df=N-1
  ciMult <- qt(conf.interval/2 + .5, datac$N-1)
  datac$ci <- datac$se * ciMult
  
  return(datac)
}

```
## Import and check data

```{r}
data <- read.table('LL_LP_SAA_cytokine_data_0925.csv', sep = ",", header = TRUE, quote = "")
  #change ID and group to factor variables
data %<>% mutate_at(c("ID", "group"), factor)
head(data)
str(data)
```

## Plot all SAA data, with trend lines for each horse
```{r}
plot1 <- ggplot(data, aes(hrs, saa))
plot1 + geom_jitter(aes(color = group)) +
  geom_smooth(aes(color = group), se = FALSE) +
  geom_line(aes(group = ID))
```

## There is one outlier, check cytokine data for similar

```{r}
  # add separator for new controls and older data
data$time <- ifelse(data$ID == 'Vesunna'|
                      data$ID == 'Rita'|
                      data$ID == 'Myla',
                                      'new', 'old') %>% factor()
str(data)  
  #test plot
labels1 <- data %>% filter(time == 'new' & hrs == 72)
labels2 <- data %>% filter(ID == 'Rita' & hrs == 48)
  # create separately d/t Rita data ending at 48 hrs
p <- ggplot(data, aes(hrs, Eotaxin)) +
  geom_line(aes(group = ID, color = time)) +
  geom_text_repel(data = labels1, aes(label = ID),
  fontface ="plain", color = "black", size = 3) +
  geom_text_repel(data = labels2, aes(label = ID),
                  fontface ="plain", color = "black", size = 3)
# works as expected
p2 <- ggplot(data, aes(hrs, G.CSF)) +
  geom_line(aes(group = ID, color = time)) +
  geom_text_repel(data = labels1, aes(label = ID),
                  fontface ="plain", color = "black", size = 3) +
  geom_text_repel(data = labels2, aes(label = ID),
                  fontface ="plain", color = "black", size = 3)
p3 <- ggplot(data, aes(hrs, IL1a)) +
  geom_line(aes(group = ID, color = time)) +
  geom_text_repel(data = labels1, aes(label = ID),
                  fontface ="plain", color = "black", size = 3) +
  geom_text_repel(data = labels2, aes(label = ID),
                  fontface ="plain", color = "black", size = 3)
p4 <- ggplot(data, aes(hrs, IL.5)) +
  geom_line(aes(group = ID, color = time)) +
  geom_text_repel(data = labels1, aes(label = ID),
                  fontface ="plain", color = "black", size = 3) +
  geom_text_repel(data = labels2, aes(label = ID),
                  fontface ="plain", color = "black", size = 3)
p5 <- ggplot(data, aes(hrs, IL.18)) +
  geom_line(aes(group = ID, color = time)) +
  geom_text_repel(data = labels1, aes(label = ID),
                  fontface ="plain", color = "black", size = 3) +
  geom_text_repel(data = labels2, aes(label = ID),
                  fontface ="plain", color = "black", size = 3)
p6 <- ggplot(data, aes(hrs, IL.6)) +
  geom_line(aes(group = ID, color = time)) +
  geom_text_repel(data = labels1, aes(label = ID),
                  fontface ="plain", color = "black", size = 3) +
  geom_text_repel(data = labels2, aes(label = ID),
                  fontface ="plain", color = "black", size = 3)
p7 <- ggplot(data, aes(hrs, IL.17A)) +
  geom_line(aes(group = ID, color = time)) +
  geom_text_repel(data = labels1, aes(label = ID),
                  fontface ="plain", color = "black", size = 3) +
  geom_text_repel(data = labels2, aes(label = ID),
                  fontface ="plain", color = "black", size = 3)
p8 <- ggplot(data, aes(hrs, IL.2)) +
  geom_line(aes(group = ID, color = time)) +
  geom_text_repel(data = labels1, aes(label = ID),
                  fontface ="plain", color = "black", size = 3) +
  geom_text_repel(data = labels2, aes(label = ID),
                  fontface ="plain", color = "black", size = 3)
p9 <- ggplot(data, aes(hrs, IL.4)) +
  geom_line(aes(group = ID, color = time)) +
  geom_text_repel(data = labels1, aes(label = ID),
                  fontface ="plain", color = "black", size = 3) +
  geom_text_repel(data = labels2, aes(label = ID),
                  fontface ="plain", color = "black", size = 3)
p10 <- ggplot(data, aes(hrs, IFNg)) +
  geom_line(aes(group = ID, color = time)) +
  geom_text_repel(data = labels1, aes(label = ID),
                  fontface ="plain", color = "black", size = 3) +
  geom_text_repel(data = labels2, aes(label = ID),
                  fontface ="plain", color = "black", size = 3)
p11 <- ggplot(data, aes(hrs, IL.8)) +
  geom_line(aes(group = ID, color = time)) +
  geom_text_repel(data = labels1, aes(label = ID),
                  fontface ="plain", color = "black", size = 3) +
  geom_text_repel(data = labels2, aes(label = ID),
                  fontface ="plain", color = "black", size = 3)
p12 <- ggplot(data, aes(hrs, IP.10)) +
  geom_line(aes(group = ID, color = time)) +
  geom_text_repel(data = labels1, aes(label = ID),
                  fontface ="plain", color = "black", size = 3) +
  geom_text_repel(data = labels2, aes(label = ID),
                  fontface ="plain", color = "black", size = 3)
p13 <- ggplot(data, aes(hrs, IL.10)) +
  geom_line(aes(group = ID, color = time)) +
  geom_text_repel(data = labels1, aes(label = ID),
                  fontface ="plain", color = "black", size = 3) +
  geom_text_repel(data = labels2, aes(label = ID),
                  fontface ="plain", color = "black", size = 3)
p14 <- ggplot(data, aes(hrs, TNFa)) +
  geom_line(aes(group = ID, color = time)) +
  geom_text_repel(data = labels1, aes(label = ID),
                  fontface ="plain", color = "black", size = 3) +
  geom_text_repel(data = labels2, aes(label = ID),
                  fontface ="plain", color = "black", size = 3)

plot_list <- list(p, p2, p3, p4, p5, p6, p7, p8, p9, p10, p11, p12, p13, p14)
combined_plot <- patchwork::wrap_plots(plot_list, nrow=4, guides='collect')
print(combined_plot)
```

## Similar, there seems to be a single outlier. Check formal test:

```{r}
# check time zero for initial outliers
saa_0 <- data %>% filter (hrs == 0) %>% .$saa
dixon.test(saa_0, two.sided=F)
  #yes, value of 1716 at time 0 is an outlier in horse AB
  #Check eotaxin, gcsf and IL8
eotaxin_0 <- data %>% filter (hrs == 0) %>% .$Eotaxin
dixon.test(eotaxin_0, two.sided = F)
gcsf_0 <- data %>% filter (hrs == 0) %>% .$G.CSF
dixon.test(gcsf_0, two.sided = F)
IL8_0 <- data %>% filter (hrs == 0) %>% .$IL.8
dixon.test(IL8_0, two.sided = F)
  #all yes, Myla is an outlier
```

## Given the above analysis, suggest removing both horses from analysis. Both
## were outliers at time 0, although interestingly, they were only outliers for one anaysis (either SAA or cytokines), but not both...

## Remove outliers and summarize SAA data

```{r}
data2 <- data %>% filter(ID != 'AB' & ID != 'Myla')

data_summary <- data2 %>% 
  group_by(hrs, group) %>%
  summarise(mean = mean(saa), sd = sd(saa), min = min(saa), max = max(saa))
#knitr::kable(data_summary)
kbl(data_summary, longtable = T, booktabs = T, align = "c") %>%
  column_spec(1, bold=T) %>%
  collapse_rows(columns = 1:2, latex_hline = "major", valign = "middle")
```

### Create model
```{r}

# create model
model_small <- lmer(saa ~ group*hrs + (1|ID), data = data2)
coef(summary(as(model_small, "lmerModLmerTest"), ddf="Kenward-Roger"))
# note the only real difference is over time, no groups are different in this analysis
```

## The only differences are in time, no groups are different overall
### Check if there are differences between groups at any time point?
```{r}
emm_small2 <- emmeans(model_small, ~ group|hrs, at = list(hrs=c(0,0.5, 1, 2, 4, 8, 24, 48, 72)))
#emm_small2
pairs(emm_small2) # there are small differences in 20 vs 40 at 48 & 72 hrs
                  # with 40 being lower than 20
```
### It looks like the control and LB40 groups are different at 48 and 72 hrs (with
###   40 being lower)

## Plot the differences

```{r}
saa_plot_data <- data %>%
  filter(ID != 'AB' & ID != 'Myla') %>%
  summarySE(measurevar = 'saa', groupvars = c('group', 'hrs'))
# set jitter
pd <- position_dodge(3)
# plot saa
saa_plot <- ggplot(data = saa_plot_data, aes(x = hrs, y = saa, group = group)) +
  geom_errorbar(aes(ymin=saa-ci, ymax=saa+ci), width=2, position=pd) +
  geom_line(aes(linetype = group), position=pd) +
  geom_point(aes(shape = group), size = 3, position=pd) +
  annotate(geom = 'text', x = c(48, 72), y = c(2450, 2350), label = '*', size = 10)
print(saa_plot)
```

## There were small differences between
## groups for Eotaxin, G-CSF and IL-8
### Eotaxin - all time points control was different from both groups except at time 72 hrs, where LB20 was not different from control
```{r}
model_eotaxin <- lmer(Eotaxin ~ group*hrs + (1|ID), data = data2)
coef(summary(as(model_eotaxin, "lmerModLmerTest"), ddf="Kenward-Roger")) # Nothing there
emm_eotaxin <- emmeans(model_eotaxin, ~ group|hrs, at = list(hrs=c(0,0.5, 1, 2, 4, 8, 24, 48, 72)))
pairs(emm_eotaxin)
contrast(emm_eotaxin, "trt.vs.ctrl")
eotaxin_plot_data <- data %>%
  filter(ID != 'AB' & ID != 'Myla') %>%
  summarySE(measurevar = 'Eotaxin', groupvars = c('group', 'hrs'))
# set jitter
pd <- position_dodge(3)
# plot
eotaxin_plot <- ggplot(data = eotaxin_plot_data, aes(x = hrs, y = Eotaxin, group = group)) +
  geom_errorbar(aes(ymin=Eotaxin-ci, ymax=Eotaxin+ci), width=2, position=pd) +
  geom_line(aes(linetype = group), position=pd) +
  geom_point(aes(shape = group), size = 3, position=pd) 
  #annotate(geom = 'text', x = c(48, 72), y = c(2450, 2350), label = '*', size = 10)
print(eotaxin_plot)
```
### G-CSF
```{r}
model_gcsf <- lmer(G.CSF ~ group*hrs + (1|ID), data = data2)
coef(summary(as(model_gcsf, "lmerModLmerTest"), ddf="Kenward-Roger")) # Nothing there
emm_gcsf <- emmeans(model_gcsf, ~ group|hrs, at = list(hrs=c(0,0.5, 1, 2, 4, 8, 24, 48, 72)))
pairs(emm_gcsf)
contrast(emm_gcsf, "trt.vs.ctrl")
gcsf_plot_data <- data %>%
  filter(ID != 'AB' & ID != 'Myla') %>%
  summarySE(measurevar = 'G.CSF', groupvars = c('group', 'hrs'))
# set jitter
pd <- position_dodge(3)
# plot
gcsf_plot <- ggplot(data = gcsf_plot_data, aes(x = hrs, y = G.CSF, group = group)) +
  geom_errorbar(aes(ymin=G.CSF-ci, ymax=G.CSF+ci), width=2, position=pd) +
  geom_line(aes(linetype = group), position=pd) +
  geom_point(aes(shape = group), size = 3, position=pd)
  #annotate(geom = 'text', x = c(48, 72), y = c(2450, 2350), label = '*', size = 10)
print(gcsf_plot)
```

### IL-8
```{r}
model_IL.8 <- lmer(IL.8 ~ group*hrs + (1|ID), data = data2)
coef(summary(as(model_IL.8, "lmerModLmerTest"), ddf="Kenward-Roger")) # 20 & 40 diff early on
emm_IL.8 <- emmeans(model_IL.8, ~ group|hrs, at = list(hrs=c(0,0.5, 1, 2, 4, 8, 24, 48, 72)))
pairs(emm_IL.8)
contrast(emm_IL.8, "trt.vs.ctrl")

IL8_plot_data <- data %>%
  filter(ID != 'AB' & ID != 'Myla') %>%
  summarySE(measurevar = 'IL.8', groupvars = c('group', 'hrs'))
# set jitter
pd <- position_dodge(3)
# plot
IL8_plot <- ggplot(data = IL8_plot_data, aes(x = hrs, y = IL.8, group = group)) +
  geom_errorbar(aes(ymin=IL.8-se, ymax=IL.8+se), width=2, position=pd) +
  geom_line(aes(linetype = group), position=pd) +
  geom_point(aes(shape = group), size = 3, position=pd) +
  annotate(geom = 'text', x = c(0, 0.5, 1, 2, 4, 8, 24), y = c(275, 300, 325, 350, 350, 350, 350), label = '*', size = 10)
print(IL8_plot)
```

## There were also differences in a few cytokines at 72 hrs that I believe were
## unrelated to the study but were in fact related to the analysis - spurious data
## These 2 plots look at the aggregate data and individual results
```{r}
IL18_plot_data <- data %>%
  filter(ID != 'AB' & ID != 'Myla') %>%
  summarySE(measurevar = 'IL.18', groupvars = c('group', 'hrs'))
# set jitter
pd <- position_dodge(3)
# plot
IL18_plot <- ggplot(data = IL18_plot_data, aes(x = hrs, y = IL.18, group = group)) +
  geom_errorbar(aes(ymin=IL.18-ci, ymax=IL.18+ci), width=2, position=pd) +
  geom_line(aes(linetype = group), position=pd) +
  geom_point(aes(shape = group), size = 3, position=pd) +
  annotate(geom = 'text', x = 72, y = 280, label = '*', size = 10)
#print(IL18_plot)
IL18_raw <- ggplot(data2, aes(hrs, IL.18))
IL18_raw_plot <- IL18_raw + geom_jitter(aes(color=group)) +
  geom_smooth(aes(color=group), se=FALSE) +
  geom_line(aes(group=ID))

IL6_plot_data <- data %>%
  filter(ID != 'AB' & ID != 'Myla') %>%
  summarySE(measurevar = 'IL.6', groupvars = c('group', 'hrs'))
# set jitter
pd <- position_dodge(3)
# plot
IL6_plot <- ggplot(data = IL6_plot_data, aes(x = hrs, y = IL.6, group = group)) +
  geom_errorbar(aes(ymin=IL.6-se, ymax=IL.6+se), width=2, position=pd) +
  geom_line(aes(linetype = group), position=pd) +
  geom_point(aes(shape = group), size = 3, position=pd) +
  annotate(geom = 'text', x = 72, y = 125, label = '*', size = 10)
#print(IL6_plot)
IL6_raw <- ggplot(data2, aes(hrs, IL.6))
IL6_raw_plot <- IL6_raw + geom_jitter(aes(color=group)) +
  geom_smooth(aes(color=group), se=FALSE) +
  geom_line(aes(group=ID))

IL2_plot_data <- data %>%
  filter(ID != 'AB' & ID != 'Myla') %>%
  summarySE(measurevar = 'IL.2', groupvars = c('group', 'hrs'))
# set jitter
pd <- position_dodge(3)
# plot
IL2_plot <- ggplot(data = IL2_plot_data, aes(x = hrs, y = IL.2, group = group)) +
  geom_errorbar(aes(ymin=IL.2-se, ymax=IL.2+se), width=2, position=pd) +
  geom_line(aes(linetype = group), position=pd) +
  geom_point(aes(shape = group), size = 3, position=pd) +
  annotate(geom = 'text', x = 72, y = 110, label = '*', size = 10)
#print(IL2_plot)
IL2_raw <- ggplot(data2, aes(hrs, IL.2))
IL2_raw_plot <- IL2_raw + geom_jitter(aes(color=group)) +
  geom_smooth(aes(color=group), se=FALSE) +
  geom_line(aes(group=ID))

IL.2.6.18_plot <- patchwork::wrap_plots(IL2_plot, IL6_plot, IL18_plot,
                                        nrow=2, guides='collect')
print(IL.2.6.18_plot)

IL.2.6.18_raw_plot <- patchwork::wrap_plots(IL2_raw_plot, IL6_raw_plot, IL18_raw_plot,
                                            nrow=2, guides='collect')
print(IL.2.6.18_raw_plot)
```


# Code used to evaluate and graph the data

```{r ref.label=knitr::all_labels(), echo=TRUE, eval=FALSE}
```
